---
title: "ascent: A multi-layer framework for decomposing functional community restructuring into positional, dispersive and boundary components under hierarchical null models."
author: "Rogelio R. Muñoz-Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ascent: A multi-layer framework for decomposing functional community restructuring into positional, dispersive and boundary components under hierarchical null models.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 10, 
  fig.height = 5.5
)
```

## Introduction

The `ascent` package implements the **ASC-FCD** (A multi-layer framework for decomposing functional community restructuring into positional, dispersive and boundary components under hierarchical null models.) framework. It provides a multi-layer approach to decompose functional community restructuring into three orthogonal geometric components:

1. **Positional Shift ($\Delta C$):** The net directional displacement of the community centroid.
2. **Dispersive Shift ($\Delta FDis$):** The internal demographic reorganization (expansion or contraction of niche variance).
3. **Boundary Shift ($\Delta FRic$):** The expansion or collapse of extreme functional phenotypes (Convex Hull volume).

To isolate deterministic environmental filtering from stochastic noise, `ascent` evaluates these shifts against a hierarchical triad of null models: Structural (Incidence), Quantitative (Demographic), and Identity (Trait Shuffle).

## 1. Simulating an Avian Metacommunity

We will simulate a bird community containing mixed data types: quantitative traits (body mass, beak length) and binary traits (diet, habitat specialization). 

We will define two abundance matrices: one to evaluate a **temporal perturbation** (deforestation) and one to evaluate a **spatial gradient** (Primary Forest $\rightarrow$ Secondary Forest $\rightarrow$ Agriculture).

```{r setup}
library(ascent)

# 1. Functional Trait Space (10 Bird Species)
# Binary traits must be factors so the Gower metric processes them correctly.
set.seed(42)
aves_traits <- data.frame(
  Body_Mass   = c(15, 20, 30, 45, 60, 90, 150, 250, 400, 600), # Quantitative (g)
  Beak_Length = c(10, 12, 15, 22, 28, 35,  45,  60,  85, 120), # Quantitative (mm)
  Frugivore   = factor(c(0, 0, 1, 1, 1, 0,  1,  1,  1,  0)),   # Binary (0/1)
  Forest_Dep  = factor(c(0, 0, 0, 0, 1, 1,  1,  1,  1,  1))    # Binary (0/1)
)
rownames(aves_traits) <- paste0("Av_sp", 1:10)

# 2a. Temporal Matrix (Deforestation favors tiny birds)
abund_time <- rbind(
  Site1_Ref = c( 0,  0, 10, 15, 25, 20, 10, 8, 4, 2),
  Site1_Imp = c(40, 35, 15,  5,  0,  0,  0, 0, 0, 0)
)
sitios <- c("Site1", "Site1")
tiempos <- c("Reference", "Impacted")

# 2b. Spatial Matrix (Agriculture favors mid-sized opportunistic birds)
abund_space <- rbind(
  Primary   = c( 0,  0,  5, 10, 20, 25, 15, 10, 5, 2),
  Secondary = c( 0,  5, 10, 20, 35, 20, 10,  0, 0, 0),
  Agri      = c( 0,  0,  0,  5, 45, 45,  5,  0, 0, 0) # Foco demográfico en Sp5 y Sp6
)
```

---

## Part I: Baseline Functional Entities (`asc_entities`)

Before comparing communities, we extract the absolute functional topology (baseline CWM, FDis, and FRic) of our spatial sites. Because we have mixed data, we set `dist_method = "gower"`.

```{r entities}
# Evaluate isolated entities
base_topo <- asc_entities(traits = aves_traits, abund = abund_space, dist_method = "gower")
summary(base_topo)
```

*Observation:* The `Primary` forest supports a massive functional volume (FRic) and high dispersion, while the `Agri` matrix is functionally poor and highly clustered around generalist traits.

---

## Part II: Temporal Functional Shifts (`asc_paired` & `asc_transitions`)

We will calculate the functional restructuring driven by deforestation in Site 1, triggering the triad of null models. *(Note: We use `n_perm = 99` for CRAN vignette compilation speed. Use 999 or 9999 for robust research).*

```{r paired}
# Execute the paired multi-layer framework
res_paired <- asc_paired(
  traits = aves_traits, abund = abund_time, sites = sitios, time = tiempos,
  ref_time = "Reference", dist_method = "gower"
)
res_paired <- asc_null(res_paired, n_perm = 99, seed = 123)
summary(res_paired)
```

### Visualizing Dynamics and Extracting Leverage

The `plot()` method projects the dynamic trajectory and the Convex Hulls. To understand *who* is driving this topological shift, we use `asc_transitions()`.

```{r paired_plot}
# Plot Trajectory (Layer 1 & 3) alongside the Leverage drivers
plot(res_paired, contrast = "Site1", type = "both", n_sp = 5)

# Extract full numerical leverage data
drivers_time <- asc_transitions(res_paired)
head(drivers_time$Site1$species_leverage, 4)
```

*Ecological meaning:* `Av_sp1` (positive leverage) actively pulled the ecosystem towards the degraded state, while `Av_sp7` (negative leverage) was a forest specialist whose extinction "released" the centroid from its mature equilibrium.

---

## Part III: Spatial Functional Networks (`asc_pairwise`)

For landscape or beta-diversity studies, `ascent` computes the bidirectional spatial divergence between all possible community pairs.

```{r pairwise}
# Execute spatial network framework
res_pw <- asc_pairwise(traits = aves_traits, abund = abund_space, dist_method = "gower")
res_pw <- asc_null(res_pw, n_perm = 99, seed = 42)

# Summarize the entire spatial functional network
summary(res_pw)
```

### Visualizing Spatial Divergence

We can plot specific spatial links within the network. Let's visualize the massive functional gap between the `Primary` forest and the `Agri` matrix.

```{r pairwise_plot}
# Plot spatial topological divergence
plot(res_pw, contrast = "Primary_vs_Agri", type = "both", n_sp = 5)
```

The resulting spatial topology clearly demonstrates that agricultural intensification does not simply result in species loss, but forces a deterministic collapse and complete directional displacement of the multidimensional functional niche.
